Synthesizing Quotient Lenses

### 2017-11-02 papers/QuotientLensSynth/main.tex
some edits and comments in sections 2 and 3

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\project}[2]{\ensuremath{\mathit{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash} \; #1 \rightarrow #2
\mathit{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}\; (#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}
\newcommand{\canonizer}{\ensuremath{\mathit{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\mathit{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\mathit{put}}}
\newcommand{\create}{\ensuremath{\mathit{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{Boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer often
has to separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these three
share a great deal of structure.  Moreover, writing the lens between the
equivalence classes can be challenging due to the rather fiddly, mainly due to
``non-ambiguity'' conditions in the lens well-formedness rules.

This paper addresses both of these issues.  First, in Section~\ref{example} we
explain the difficulties of programming with quotient lenses and show, by way of
example, how our novel language of {\em Quotient Regular Expressions}, or QREs
and QRE lenses address these difficulties.  Quotient regular expressions allow
programmers to specify a broad class of equivalence relations over regular
languages using a compact and convenient notation in which the regular language
and its equivalence classes are captured by a single expression.  We then
define {\em QRE lenses}, a class of bijective quotient lenses that are typed
using QREs, and we show how to {\em synthesize} QRE lenses from their QRE types
and a set of input-output examples.  We demonstrate the practicality of this
approach by synthesizing QRE lenses between several real-world data formats
from the {\tt data.gov} database.

@@

\section{Quotient Regular Expressions and Lenses by Example}
Consider a bidirectional transformation which converts citation records such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

For example, with these rules, the following \bibtex{} citations should be
considered the same by the lens,
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }

    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "Conway, J. H.",     }
\end{verbatim}
with the EndNote records having similar rules defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences which nest and interact
with each other in a way that makes them difficult to express with the standard
Boomerang combinators. Moreover, many of the equivalences expressible in
Boomerang require that the programmer first specify the data itself in form of
a regular expression before specifying the equivalence relation on the data.
This leads to verbose, repetitive code.

To address this issue, we propose {\em Quotient Regular Expressions} or QREs,
which express a broad class of equivalence relations directly on regular
languages using a compact and convenient notation. For example, with QREs we
can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]
let wsp_to_space = project wsp+ $\to$ " "
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)
let comma_name = name | (name . "," . wsp_to_space . (name . wsp_to_space)* . name)
let space_to_comma = synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")} 
let valid_name = squash comma_name -> space_name using (get space_to_comma)
let wsp_to_and = project (wsp+ . "and" . wsp+) $\to$ " and " 
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name) 
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word) 
let bib_title = "title = {" . title . "}" 
let bib_perm = perm (label, bib_author, bib_title) with (project ("," .wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

The expression \cd{project [ \\n\\t\\r]}$^+$ $\to$ \cd{" "} is a QRE that
matches any number of whitespace characters, and then considers all strings of
whitespace characters as equivalent with a single space as the canonical
representative. Another QRE in the example above is the expression

\begin{center}
\begin{lstlisting}
squash comma_name -> space_name using (get space_to_comma)
\end{lstlisting}
\end{center}

\noindent which considers the \cd{comma_name} representation of the author's
name to be equivalent to the \cd{space_name} representation, with \cd{get
space_to_comma} giving the function that transforms names in the \cd{comma_name}
representation to names in the \cd{space_name} representation.

The third example of a QRE in the example is the expression
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent QRE, which specifies that the label, title and author fields can occur
in any order, with the fields separated by a comma followed by arbitrary amounts
of whitespace. The canonical representation is that in which the label, author
and title fields occur in that order, separated by a comma followed a newline.
The \cd{perm} QRE allows for easy nesting of permutations. Similarly, it is easy
to specify any number of fields within the \cd{perm} QRE.

Now QREs make it easier to write quotient lenses by specifying the data and
the equivalence relation on the data in one expression, but the programmer still
has to manually derive the underlying lens which maps the equivalence classes
of the source data to the equivalence classes of the target data. This is
itself a tedious and error-prone task, as the code snippet below illustrates:

\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") = 
    del "@book{" . ins "%0 Book\n%F "

@@

let bib_to_end : (lens in bib_tex $\Leftrightarrow$ end_note) =
   preamble . label . author_lens . title_lens
\end{lstlisting}

This lens transforms the \bibtex{} data to the EndNote (and vice versa) data by
matching the delimiters and separators in a bijective manner. For example,
the \cd{preamble} lens replaces the \bibtex{} preamble \cd{@book} with the
EndNote preamble \cd{\%0 Book\n\%F } in the \bibtex{} record. The programmer
can ill afford to make a mistake with any one of the small but vital details
of the transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

Therefore, to address this challenge, we identify a special class of quotient
lenses which we call {\em QRE lenses}, and show how to {\em
synthesize} QRE lenses from a pair of QREs and a set of input-output
example pairs. Our key insight is to reduce the problem of synthesizing QRE
lenses to the problem of synthesizing {\em bijective lenses}, which are the
bijections created from the Boomerang lens combinators. This is a problem
which we solved in our prior work ~\cite{popl18}.

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs - every QRE
lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

@@

One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Our main technical contribution in this paper is a
proof that each QRE lenses $q$ defines the same transformation as a QRE lens
$q'$ which is in normal form.

For example, suppose \cd{q_1 : bibtex <=> endnote} is a QRE lens that maps
between \bibtex{} and EndNote formats, and that \cd{q_2 : endnote <=> xml}
is a QRE lens that maps between EndNote and XML formats. The composition 
\cd{(q_1 ; q2) : bibtex <=> xml} of \cd{q_1} and \cd{q_2} is not in normal form
due to the presence of canonizers on each side of the intermediary EndNote
format:

@@

\noindent However, while \cd{q_1 ; q_2} is not in normal form, there is another
QRE lens \cd{q : bibtex <=> xml} of the same type as \cd{q_1 ; q_2} such that
\cd{q_1 ; q_2} and \cd{q} define the same transformation. This proof will allow
justify our approach for synthesizing quotient lenses so that the programmer
can simply type

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnote) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent to synthesize a quotient lens that maps between the \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_bexample} to
the equivalence class of \cd{end_example}.

@@

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep \project{R}{s} \sep \squash{q}{q'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \;  | \; \normalize{R}{R'}{f}\\
&q \semicolon q' \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

@@

\begin{lstlisting}
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)
\end{lstlisting}

The $q \semicolon q'$ QRE is a compostion combinator used to successively apply
two equivalence relations to a regular language. For example, in the \bibtex{}
to EndNote example, when defining the \cd{valid_name} QRE which maps
\cd{comma_name} into \cd{space_name}, we first defined the equivalence
relations on the \cd{comma_name} and \cd{space_name} representations before
defining the \cd{valid_name} QRE. Using the composition combinator, we could
instead have defined \cd{valid_name} as follows:

@@

The final combinator is the  $\normalize{R}{R'}{f}$ combinator, which expresses
the equivalence relation which sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QREs
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations which are
difficult to express, or which cannot be expressed, using the other
combinators.

However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

While QREs express a broad class of equivalence relations on regular languages,
they do not express all the equivalence relations that can be expressed in
Boomerang, nor of course all the equivalence relations which cannot be
expressed using Boomerang canonizers. For instance, Boomerang canonizers can
map a regular language described by $R$ to a regular language described by
$S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $q$
canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$.

In summary each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $\eqrel{q}$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $\eqrel{q}$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{q}$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $\eqrel{q}$,
  then the equivalence classes of $\eqrel{q}$ are the same as the fibres of
  $\canonizer(q)$.

Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity} which comes up when defining QREs. If $q, q'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(q) \cdot W(q')$, $W(q) \sep W(q')$,
$W(q)^*$, $K(q) \cdot K(q')$, $K(q) \sep K(q')$, $K(q)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.

@@

The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $\eqrel{q}$ and the
canonizing function $\canonizer(q)$ are given in
Figures~\ref{fig:wk},~\ref{fig:relations} and~\ref{fig:canonizers}.
\begin{figure}[t]
  \centering
  \[
    \begin{array}{l@{\quad}l@{\quad}l}

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2
\kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{Boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer often
has to separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these three
share a great deal of structure.  Moreover, writing the lens between the
equivalence classes can be challenging due to the rather fiddly typechecking, mainly due to
``non-ambiguity'' conditions in the lens well-formedness rules.

This paper addresses both of these issues.  First, in Section~\ref{sec:example} we
explain the difficulties of programming with quotient lenses and show, by way of
example, how our novel language of {\em Quotient Regular Expressions}, or QREs
and QRE lenses address these difficulties.  Quotient regular expressions allow
programmers to specify a broad class of equivalence relations over regular
languages using a compact and convenient notation in which the regular language
and its equivalence classes are captured by a single expression.  We then
define {\em QRE lenses}, a class of bijective quotient lenses that are typed
using QREs, and we show how to {\em synthesize} QRE lenses from their QRE types
and a set of input-output examples.  We demonstrate the practicality of this
approach by synthesizing QRE lenses between several real-world data formats
from the {\tt data.gov} database.

@@

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}

Consider a bidirectional transformation which converts citation records such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

For example, with these rules, the following \bibtex{} citations should be
considered the same by the lens \saz{``the lens'' hasn't been introduced to the
  reader yet},
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }

    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "Conway, J. H.",     }
\end{verbatim}
with the EndNote records having similar rules \saz{similarly, ``rules'' haven't been
  mentioned yet, do you mean similar ``equivalence''} defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences \saz{explicitly mention
  whitespace and permutation?} which nest and interact
with each other in a way that makes them difficult to express with the standard
Boomerang combinators. Moreover, many of the equivalences expressible in
Boomerang require that the programmer first specify the data itself in form of
a regular expression before specifying the equivalence relation on the data.
\saz{Something like this instead?:  Moreover, Boomerang requires that the
  programmer first specify the data itself in the form of a regular expression
  before she can define the equivalence relation on that data.}
This leads to verbose, repetitive code.

To address this issue\saz{these issues? these problems with Boomerang?}, we propose {\em Quotient Regular Expressions} or QREs,
which express a broad class of equivalence relations directly on regular
languages using a compact and convenient notation. For example, with QREs we
can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]
let wsp_to_space   = project wsp$^+$ $\to$ " "
let space_name     = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let comma_name   = name | (name . "," . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let space_to_comma = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")} 
let valid_name = squash comma_name $\to$ space_name using (get space_to_comma)
let wsp_to_and = project (wsp$^+$ . "and" . wsp$^+$) $\to$ " and " 
let bib_names  = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name) 
let bib_author = "author = \"" . bib_names . "\""
let title     = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word) 
let bib_title = "title = {" . title . "}" 
let bib_perm  = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex    = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

\saz{Throughout these examples, refer to the definitions in Figure 1.}  In the
figure, the QRE \cd{wsp_to_space}, which is defined to be \cd{project wsp}$^+$
$\to$ \cd{" "}, matches any non-zero number of whitespace characters, but
considers the matched string of whitespace characters as equivalent with a
single space as the canonical representative.  \saz{We call the type of QRE a
\textit{projection} because it projects an entire regular language down to a single
representative string in the language.}

\saz{
Another form of QRE is the \textit{squash} QRE, an example of which is given by
\cd{valid_name}, which is defined to be:
}

\begin{center}
\begin{lstlisting}
squash comma_name $\to$ space_name using (get space_to_comma)
\end{lstlisting}
\end{center}

\noindent This QRE specifies that the \cd{comma_name} representation of the
author's name is equivalent to the \cd{space_name} representation.  Here \cd{get
  space_to_comma} is the function that transforms names in the \cd{comma_name}
representation to names in the \cd{space_name} representation.  \saz{In this
  example, \cd{space_to_comma} is itself an automatically \textit{synthesized}
  lens between the \cd{space_name} and \cd{comma_name} QREs, and we need to use
  the ``get'' direction of the lens to specify the squash transformation. (We'll
  say more about synthesis below.)}

The third example of a QRE in the Figure is \cd{bib_perm}, defined as:
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent This is a \textit{permutation} QRE, which specifies that the label, title, and author fields can occur
in any order, with the fields separated by a comma followed by arbitrary amounts
of whitespace. The canonical representation is that in which the label, author
and title fields occur in that order, separated by a comma followed a newline.
Such \cd{perm} QREs allow for easy nesting of permutations. Similarly, it is easy
to specify any number of fields within the \cd{perm} QRE.

\saz{Given the \textit{projection}, \textit{squash}, and \textit{permutation}
  QRE primitives, we can build more complex languages using the usual regular
  expression combinators of alternative choice `\cd{|}', Kleene star `$^*$', and
  sequential composition ``.''.  As shown in the rest of
  Figure~\ref{fig:example}, it is straightforward to define both a regular
  language that accepts \bibtex{} records \textit{and}, simultaneously, an
  equivalence relation that quotients out irrelevant details.}

\subsection{QRE Lenses}
QREs make it easier to write quotient lenses by specifying the data and the
equivalence relation on the data in one expression, but to use them to define a
lens, the programmer still has to manually derive the underlying lens that maps
the equivalence classes of the source data to the equivalence classes of the
target data.

\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") = 
    del "@book{" . ins "%0 Book\n%F "

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) =
   preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example lens between the \cd{bibtex} and \cd{end_note} QRE languages.}
\label{fig:example-lens}
\end{figure}

This is itself a tedious and potentially error-prone task, as the code snippet
in Figure~\ref{fig:example-lens} illustrates.  The lens shown in the Figure
transforms \bibtex{} data to EndNote data  (and vice versa) by matching
the delimiters and separators in a bijective manner. In more detail, the
\cd{preamble} lens replaces the \bibtex{} preamble \verb|@Book| with the EndNote
preamble \verb|%0 Book\n%F|. \saz{This next sentence isn't very well-supported
  by the previous sentence's example, since that doesn't seem too hard.  Maybe
  explain some of the harder parts too?}  The programmer can ill
afford to make a mistake with any one of the small but vital details of the
transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

To address this challenge, we identify a \saz{avoid ``special''} class of quotient
lenses that \saz{instead of ``which''} we call {\em QRE lenses}, and show how to {\em
synthesize} QRE lenses from a pair of QREs and a set of input-output
example pairs. Our key insight is to reduce the problem of synthesizing QRE
lenses to the problem of synthesizing {\em bijective lenses}, which are the
bijections created from the Boomerang lens combinators. This is a problem
which we solved in our prior work ~\cite{popl18}.

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs---\saz{in
  latex use three dashes --- for this (not one -).  Put no spaces around ``---''}every QRE
lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

@@

One potential pitfall with QRE lenses is that the combinators that \saz{instead
  of ``which''}
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ might not be equivalent to
any quotient lens that \saz{instead of ``which''} has the canonical normal form of a bijective lens with
canonizers at the edges. Our main technical contribution in this paper is a
proof that each QRE lenses $q$ defines the same transformation as a QRE lens
$q'$ which is in normal form.

For example, suppose \cd{q_1 : bibtex} $\Leftrightarrow$ \cd{endnote} is a QRE
lens that maps between \bibtex{} and EndNote formats, and that \cd{q_2 : endnote
  <=> xml} is a QRE lens that maps between EndNote and XML formats. The
composition \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} of \cd{q_1} and
\cd{q_2} is not (necessarily) in normal form due to the presence of canonizers
on each side of the intermediary EndNote format:

@@

\noindent However, while \cd{q_1 ; q_2} is not in normal form, our main theorem
shows that there is another QRE lens \cd{q : bibtex} $\Leftrightarrow$ \cd{xml} of the same type as
\cd{q_1 ; q_2} such that \cd{q_1 ; q_2} and \cd{q} define the same
transformation. This proof will allow justify our approach for synthesizing
quotient lenses so that the programmer can \saz{replace all of the code in
  Figure~\ref{fig:example-lens} with a call to the synthesis prodedure:}

\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) 
= synth bibtex $\Leftrightarrow$ endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the resulting  quotient lens maps between the \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_example} to
the equivalence class of \cd{end_example}, which are the two concrete example
strings given at the beginning of this Section.

\saz{
As we saw with the definition of \cd{space_to_comma} in
Figure~\ref{fig:example-qre}, the synthesis procedure itself can be used to
creates lenses that are in turn used to define other QREs.  This ability to mix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.}

@@

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:  \saz{Use the same font that we use above---I added a \textbackslash kw
macro}
\begin{align*}
q := \; &R \sep \project{R}{s} \sep \squash{q}{q'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \\
  & | \; \normalize{R}{R'}{f} \sep q \semicolon q' \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

\saz{We use the notation $\mathcal{L}(R)$ to mean the language accepted by the
  regular expression $R$.}

@@

\begin{lstlisting}
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
\end{lstlisting}

\saz{One combinator that we haven't yet introduced is sequential composition.}
The $q \semicolon q'$ QRE successively applis
two equivalence relations to a regular language. For example, in the \bibtex{}
to EndNote example, when defining the \cd{valid_name} QRE which maps
\cd{comma_name} into \cd{space_name}, we first defined the equivalence
relations on the \cd{comma_name} and \cd{space_name} representations before
defining the \cd{valid_name} QRE. Using the composition combinator, we could
instead have defined \cd{valid_name} as follows:
\saz{Instead of replaying the \cd{valid_name} example, perhaps use a simpler
  one---even an artificial example might be better.}

@@

The final combinator is  $\normalize{R}{R'}{f}$, which expresses
the equivalence relation that \saz{instead of ``which''} sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QRE
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are
difficult to express, or that cannot be expressed, using the other
combinators.

However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable in general, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

\saz{This is a bit confusing.} \saz{The equivalence relations definable by
  QREs are a subset of those that can be expressed in Boomerang.} 
While QREs express a broad class of equivalence relations on regular languages,
they do not express all the equivalence relations that can be expressed in
Boomerang, nor of course all the equivalence relations which cannot be
expressed using Boomerang canonizers. For instance, Boomerang canonizers can
map a regular language described by $R$ to a regular language described by
$S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $q$
canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$.

The net effect of these design decisions means that each QRE $q$ enables us to
simultaneously express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $\eqrel{q}$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $\eqrel{q}$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{q}$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $\eqrel{q}$,
  then the equivalence classes of $\eqrel{q}$ are the same as the fibres of
  $\canonizer(q)$.

\subsection{QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity} which comes up when defining QREs. If $q$ and  $q'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(q) \cdot W(q')$, $W(q) \sep W(q')$,
$W(q)^*$, $K(q) \cdot K(q')$, $K(q) \sep K(q')$, $K(q)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.

@@

\subsection{Semantics of QREs}

\saz{This needs a bit more setting up / explanation. Explain what the reader
  should expect to see in the figures.}
The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $\eqrel{q}$ and the
canonizing function $\canonizer(q)$ are given in
Figures~\ref{fig:wk},~\ref{fig:relations} and~\ref{fig:canonizers}.
\begin{figure}[t]
\saz{composition notation}
  \centering
  \[
    \begin{array}{l@{\quad}l@{\quad}l}
